Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-17 manuscript_agu_rev1.tex
Update on Overleaf.

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, the model relies entirely on explicit simulation of clouds instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with the model and reanalyses. We subsetted our results by cyclonic activity and stability. We found that the model and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or near-surface clouds are especially underestimated in the reanalyses. However, the model and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat--CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

The Southern Ocean (SO) is known to be a problematic region for climate model biases \cite{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of surface and in situ observations and being a lower priority region for numerical weather prediction (NWP) and climate model development because of its distance from populated areas. Nevertheless, radiation biases and changes over an area of its size have a substantial influence on the global climate \cite{rintoul2011}, such as affecting the Earth's radiation balance, ocean heat, and carbon uptake \cite{williams2023}, and the SO is also an important part of the global ocean conveyor belt \cite{wang2014b}. In general, marine clouds have a disproportionate effect on top-of-atmosphere (TOA) SW radiation due to the relatively low albedo of the sea surface. The relative longitudinal symmetry of the SO means that model cloud biases tend to be similar across longitudes.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small a cloud fraction and too large a cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented CALIPSO Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5~km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over the high-latitude SO and an underestimation of the total cloud fraction on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48~hours of km-scale ICON limited-area model NWP simulations over an SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

@@

How attenuation of the lidar signal affects cloud detection is dependent on factors such as the optical thickness of the measured cloud and its backscattering phase function, as well as the range-dependent noise standard deviation. A rough estimate can be made under an assumption of a relatively strongly backscattering cloud of $\beta = \mathrm{100\times 10^{-6} m^{-1}sr^{-1}}$ at a height of $r_1$ = 2 km, range-dependent noise $\beta_n$ at $r_2$ = 8 km of about $\mathrm{5\times 10^{-6} m^{-1}sr^{-1}}$, and cloud detection threshold $\beta_t = \mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$, noise multiplication factor $f$ = 5, at full attenuation the two-way attenuation factor $A$ satisfies $A\beta = \beta_t + f\times \beta_n\left(\frac{r_1}{r_2}\right)^2$, which is equivalent to exponential decay ($A = e^{-2\delta}$) with optical depth $\delta$ (at the lidar wavelength) of about 1.7.

@@

When subsetted by latitude (Fig.~\ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone (40--55°S) displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone (between 55°S and the Antarctic coast), and this could be because higher latitudes have greater prevalence of weakly stable profiles (Fig.~\ref{fig:cyclone-stability}c, d), although more stable profiles populate regions south of 65°S close to the Antarctic coast. Cyclonic activity is also stronger in high-latitude SO, which is typically associated with shallow or deep convection rather than very stable stratification necessary for fog formation. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (near-surface cloud occurrence is still strongly underestimated).

@@

When subsetted by conditions of strong and weak stability (Fig.~\ref{fig:cloud-occurrence}f, g), as defined in Section~\ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly near the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1~km. Physically, conditions of strong stability are associated with the formation of advection fog, such as in situations of warm air advection from the north over a colder sea surface, thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak up to 12\%, underestimated cloud occurrence above 1~km by up to 5\%, and underestimated cloud occurrence near the surface by about 10\% in the reanalyses but not ICON. In situations of weak stability, the bias in ICON is very pronounced, with a much larger peak in cloud occurrence at about 500~m; the reanalyses underestimate cloud occurrence below 1~km, especially near the surface; and MERRA-2 underestimates cloud occurrence more strongly at almost all heights.

@@

We also analyzed the daily cloud cover (total cloud fraction) distribution. This is a measure of cloudiness, irrespective of height, calculated over the course of a day (UTC). A cloud detected at any height means that the lidar profile was classified as cloudy; otherwise, it was classified as a clear sky. When all profiles in a day are taken together, the cloud cover for the day is defined as the fraction of cloudy profiles in the total number of profiles, expressed in oktas (multiples of 1/8). The same calculation is done for the lidar observations as for the simulated lidar profiles. We use the term ``okta'' independently of its use in instantaneous synoptic observations, and here it simply means 1/8 (0.125) of the daily cloud cover.

@@

We note that the radiative transfer calculations used in the lidar simulator mean that the impact of both cloud phase and cloud fraction are convolved to produce the cloud mask. Therefore, the cloud occurrence is not affected by any cloud phase biases as long as the cloud is optically thick enough to be detected and the laser signal is not too attenuated. A combination of underestimated total cloud fraction and overestimated outgoing SW at TOA is indicative of an overestimated cloud albedo (in cloudy areas) due to either cloud liquid and ice water content, cloud phase, droplet or ice crystal size distribution, shape or orientation of ice crystals, cloud overlap, or their combination. The influence of cold clouds is likely second-order due to the much larger typical effective radius of ice crystals than cloud droplets.

In contrast to SW radiation, the models have much smaller LW radiation biases, which is expected due to the prevailing low-level clouds having similar temperatures as the surface. \citeA{roh2021} also found LW biases to be much lower than SW biases in DYAMOND models over the tropical Atlantic Ocean. In ICON, the outgoing LW radiation is overestimated by 5\% (Fig.~\ref{fig:cloud-occurrence}a). This is likely caused by an underestimated total cloud fraction exposing a larger sea surface area to cooling to space, which is typically warmer than the atmospheric temperature at 0--2~km, where most of the clouds are located. In the MERRA-2 and ERA5 reanalyses, the LW biases are also slightly positive, 4 and 5~Wm$^{-2}$, respectively. This is again in line with the underestimated total cloud fraction by about 20\%. However, if the clouds are too thick, as expected from the SW results, this might also provide a compensating effect, in which too small a cloud area is counteracted by greater optical thickness in the LW spectrum, thus reducing the outgoing LW radiation more relative to thinner clouds. For thin clouds, the outgoing TOA LW radiation originates both from the warmer surface (partly blocked by the clouds) and the clouds, whereas for thick clouds, the outgoing TOA LW radiation originates mostly from the colder-than-surface clouds.

In all the subsets (Fig.~\ref{fig:cloud-occurrence}b--g), the same type of biases are observed, namely the outgoing SW radiation is underestimated in ICON and ERA5 and overestimated in MERRA-2, and the outgoing LW radiation is overestimated in all the models. Even though the total cloud fraction is higher by 6\% over the high-latitude SO than the low-latitude SO, the outgoing SW radiation is much greater by 41~Wm$^{-2}$, implying a much greater cloud albedo (of cloudy areas) over the high-latitude SO. ICON has little difference in the total cloud fraction between low- and high-latitude SO, but greater outgoing SW radiation by 14~Wm$^{-2}$ over the high-latitude SO, likely due to thicker clouds under deeper convection in less stable and more cyclonic conditions relative to the low-latitude SO. In contrast, the reanalyses showed both greater total cloud fraction and outgoing SW radiation over the high-latitude SO compared to the low-latitude SO.

@@

In order to examine the potential link in the cloud biases to the local physical conditions, we analyzed about 2300 radiosonde profiles south of 40°S available from the 24 RV \emph{Polarstern} voyages and the MARCUS, NBP1704, TAN1702, and TAN1802 campaigns. Spatially and temporally colocated profiles were taken from ICON and the reanalyses. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with the model, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. The profiles were partitioned into the same subsets as above (Sections~\ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). We focus on comparing virtual potential temperature ($\theta_v$) and relative humidity, being one of the primary factors affecting shallow convection and the associated low-level cloud formation and dissipation. The observed and model profiles of virtual potential temperature and relative humidity are shown in Fig.~\ref{fig:potential-temperature}.

Overall, the mean $\theta_v$ is accurate to within 0.5~K in ICON and MERRA-2, except for ICON being colder by up to 2.5~K in the mid-to-high troposphere (less stable) (Fig.~\ref{fig:potential-temperature}a). Larger differences exist, however, in the 40--55°S zone, where ICON is colder by about 5 K at higher altitudes (Fig.~\ref{fig:potential-temperature}b). In other subsets, the bias is relatively small. MERRA-2 and ERA5 are very close to the observations, possibly due to a high accuracy of assimilation of this quantity. Notably, the variability of virtual potential temperature (as represented by the percentiles) is much smaller in ICON than in the observations. This indicates that the model's internal variability in the lower-tropospheric thermodynamic conditions in the SO is smaller than in reality.

@@

Fig.~\ref{fig:rs-hur-hist} shows RH histograms calculated from the radiosonde observations and equivalent profiles in the models (shown as anomalies relative to the observations), calculated for all, clear, and cloudy bins, based on the lidar observations and the simulated lidar backscatter in the models. Here, we show only the first 2 km to concentrate on the identified cloud biases seen at these heights. We can see several notable features. The models simulate progressively fewer high-RH (\textgreater 90\%) bins above the ground (Fig.~\ref{fig:rs-hur-hist}b--d). This can be related to either ice nucleation happening in the models, which requires smaller RH for saturation, or the grid cell size in the models, which requires lower grid cell average RH than 100\% for saturation to occur in a fraction of the grid cell. The models also tend to simulate more clear bins than observations for RH between 80 and 100\% between the ground and about 1 km (Fig.~\ref{fig:rs-hur-hist}f--h). In observations, these values of RH are associated with cloudy bins (Fig.~\ref{fig:rs-hur-hist}i). Conversely, the models predominantly associate only RH very close to 100\% with cloudy bins at these heights (Fig.~\ref{fig:rs-hur-hist}j--l). This may be one of the main reasons for the identified cloud or fog biases near the ground. A possible explanation is that cloud droplets are able to form or persist at RH between 90 and 100\% at these heights over the SO. This could be due to abundant sea salt or biogenic aerosol or droplet generation from sea spray in the common high swell and high wind speed conditions over the SO. Stratus fractus or other broken clouds could also lead to less than 100\% RH when averaged over the size of the vertical bins (up to 30 m in some of the radiosonde profiles).

@@

Let us consider the main limitations of the presented results. The spatial coverage of our dataset does not include most parts of the Indian Ocean and Pacific Ocean sectors of the SO. Even though climatological features of the SO are typically relatively uniform zonally, variations exist, such as those related to the Antarctic Peninsula and the southern tip of South America. The voyages were mostly undertaken in the Austral summer months and only rarely in the winter months, due to the poor accessibility of this region during winter. Therefore, our results are likely representative of summer and, to a lesser extent, spring and autumn conditions. Cloud regimes and phase in the region are seasonally variable \cite{danker2022}.

@@

Supercooled liquid clouds (liquid clouds under subzero temperature) commonly occur over the SO. In our analysis of LWP and IWP, we see that both phases are abundant. Because liquid water droplets are typically smaller and more numerous than ice crystals in cold clouds, they attenuate a greater amount of the lidar radiation. Clouds with a relatively modest optical thickness of 1.7 can attenuate the lidar signal for a detection at 2 km using an instrument with noise properties like the Vaisala CL31 (Section 2.4). While supercooled liquid clouds and their attenuation are accounted for by the lidar simulator, they can strongly attenuate the signal and cause artifically low values of cloud occurrence at higher altitudes. For example, we found that cloud occurrence at 1.5 km is underestimated in ICON and underlying clouds are overestimated. However, this can also mean that clouds at 1.5 km are present in the model but the signal is too attenuated by the lower clouds in the model but not in the observations, where the underlying clouds are not as pronounced.

>>>>

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds, temperature, and humidity profiles over the Southern Ocean in one such model, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, the model relies entirely on explicit simulation of cloud dynamics instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with the model and reanalyses. We subsetted our results by cyclonic activity and stability. We found that the model and reanalyses underestimate lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or near-surface clouds are especially underestimated in the reanalyses. However, the model and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level and lower stability. Southern Ocean cloud and fog biases are an important problem in the analyzed models and result in radiation balance biases.

@@

The main aim of this study is to evaluate the GSRM version of ICON developed by the nextGEMS project \cite{nextgems,segura2025}. ICON is developed and maintained jointly by Deutscher Wetterdienst, the Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Our aim is to quantify how well the GSRM ICON simulates clouds over the Southern Ocean (SO), particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on the Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation (CALIPSO), neither of which are now operational, are affected by the above-mentioned issues, resulting in a strong underestimation of cloud occurrence below 2 km in a merged CloudSat--CALIPSO product relative to ground-based lidar observations at McMurdo Station \cite{mcerlich2021}. Removing situations with higher overlapping clouds could enable a less biased comparison of low clouds. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and a station as a reference for the model evaluation. Altogether, we analyzed approximately 2400 days of data from 31 voyages and one sub-Antarctic station covering diverse longitudes and latitudes of the SO. To achieve a like-for-like comparison with the model, we used a ground-based lidar simulator called the Automatic Lidar and Ceilometer Framework [ALCF; \citeA{kuma2021}]. We contrasted the results with ERA5 \cite{era5} and the Modern-Era Retrospective analysis for Research and Applications, Version~2 [MERRA-2; \citeA{gelaro2017}].

@@

The SO is known to be a problematic region for climate model biases \cite{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of surface and in situ observations. This region has also long been a lower priority region for numerical weather prediction (NWP) and climate model development because of its distance from populated areas. Nevertheless, radiation biases and changes over an area of its size have a substantial influence on the global climate \cite{rintoul2011}, such as affecting the Earth's radiation balance, ocean heat, and carbon uptake \cite{williams2023}, and the SO is also an important part of the global ocean conveyor belt \cite{wang2014b}. In general, marine clouds have a disproportionate effect on top-of-atmosphere (TOA) SW radiation due to the relatively low albedo of the sea surface. The relative longitudinal symmetry of the SO means that model cloud biases tend to be similar across longitudes.

@@

SO radiation biases have been relatively large and systematic compared to the rest of the globe since at least CMIP3 \cite{trenberth2010,bodas-salcedo2012}, and the SO SW cloud radiative effect (CRE) bias is still positive in eight CMIP6 models analyzed by \citeA{schuddeboom2021} over the high-latitude SO, whereas over the low-latitude SO it tends to be more neutral or negative in some models. Too much absorbed SW radiation over the SO was also identified in the GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the ``too few too bright'' cloud bias, characterized by too small a cloud fraction and too large a cloud albedo \cite{wall2017,kuma2020}, previously described by \citeA{webb2001}, \citeA{weare2004}, \citeA{zhang2005}, \citeA{karlsson2008}, \citeA{nam2012}, \citeA{klein2013}, and \citeA{bender2017} in other regions and models, which means that a model can maintain a reasonable SW radiation balance by reflecting too much SW radiation from clouds, but these cover too small an area. A study by \citeA{konsta2022} showed that this type of bias is still present in six analyzed CMIP6 models in tropical marine clouds, using the GCM-Oriented CALIPSO Cloud Product [CALIPSO--GOCCP; \citeA{chepfer2010}] and Polarization \& Anisotropy of Reflectances for Atmospheric Sciences coupled with Observations from a Lidar [PARASOL; \citeA{lier2008}] as a reference. They suggest improper simulation of subgrid-scale cloud heterogeneity as a cause. Compensating cloud biases in the Australian Community Climate and Earth System Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were analyzed by \citeA{fiddes2022} and \citeA{fiddes2024}. \citeA{possner2022} showed that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction on the order of 30\% and overestimates net downward TOA SW radiation by approximately 10 Wm$^\mathrm{-2}$ in the highest model resolution run (2.5~km). \citeA{zhao2022} reported a similar SW radiation bias in five analyzed CMIP6 models over the high-latitude SO and an underestimation of the total cloud fraction on the order of 10\% over the entire 40--60°S SO. Recently, \citeA{ramadoss2024} analyzed 48~hours of km-scale ICON limited-area model NWP simulations over an SO region adjacent to Tasmania against the Clouds, Aerosols, Precipitation, Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN) voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found the ICON cloud optical thickness was underestimated relative to Himawari‐8 satellite observations but also identified large differences in cloud top phase.

@@

How attenuation of the lidar signal affects cloud detection is dependent on factors such as the optical thickness of the measured cloud and its backscattering phase function, as well as the range-dependent noise standard deviation \cite{kuma2021}. A rough estimate can be made under an assumption of a relatively strongly backscattering cloud of $\beta = \mathrm{100\times 10^{-6} m^{-1}sr^{-1}}$ at a height of $r_1$ = 2 km, range-dependent noise $\beta_n$ at $r_2$ = 8 km of about $\mathrm{5\times 10^{-6} m^{-1}sr^{-1}}$, and cloud detection threshold $\beta_t = \mathrm{2\times 10^{-6} m^{-1}sr^{-1}}$, noise multiplication factor $f$ = 5, at full attenuation the two-way attenuation factor $A$ satisfies $A\beta = \beta_t + f\times \beta_n\left(\frac{r_1}{r_2}\right)^2$, which is equivalent to exponential decay ($A = e^{-2\delta}$) with optical depth $\delta$ (at the lidar wavelength) of about 1.7.

@@

When the data are subset by latitude (Fig.~\ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone (40--55°S) displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone (between 55°S and the Antarctic coast), and this could be because higher latitudes have greater prevalence of weakly stable profiles (Fig.~\ref{fig:cyclone-stability}c, d), although more stable profiles populate regions south of 65°S close to the Antarctic coast. Cyclonic activity is also stronger in high-latitude SO, which is typically associated with shallow or deep convection rather than very stable stratification necessary for fog formation. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (near-surface cloud occurrence is still strongly underestimated).

@@

When the data are subset by stability (Fig.~\ref{fig:cloud-occurrence}f, g), as defined in Section~\ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly near the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1~km. Physically, conditions of strong stability are associated with the formation of advection fog, such as in situations of warm air advection from the north over a colder sea surface, thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak up to 12\%, underestimated cloud occurrence above 1~km by up to 5\%, and underestimated cloud occurrence near the surface by about 10\% in the reanalyses but not ICON. In situations of weak stability, the bias in ICON is very pronounced, with a much larger peak in cloud occurrence at about 500~m; the reanalyses underestimate cloud occurrence below 1~km, especially near the surface; and MERRA-2 underestimates cloud occurrence more strongly at almost all heights.

@@

We also analyzed the daily cloud cover (total cloud fraction) distribution. This is a measure of cloudiness, irrespective of height, calculated over the course of a day (UTC). A cloud detected at any height means that the lidar profile was classified as cloudy; otherwise, it was classified as a clear sky. When all profiles in a day are taken together, the cloud cover for the day is defined as the fraction of cloudy profiles in the total number of profiles. It is expressed in oktas (multiples of 1/8), reflecting the 3-hourly model output of MERRA-2 and ICON, i.e. 8 times per day. The same calculation is done for the lidar observations as for the simulated lidar profiles. We use the term ``okta'' independently of its use in instantaneous synoptic observations, and here it simply means 1/8 (0.125) of the daily cloud cover.

@@

We note that the radiative transfer calculations used in the lidar simulator mean that the impact of both cloud phase and cloud fraction are convolved to produce the cloud mask. Therefore, the cloud occurrence is not affected by any cloud phase biases as long as the cloud is optically thick enough to be detected and the laser signal is not too attenuated. An overestimated cloud albedo (in cloudy areas) can be due to either cloud liquid and ice water content, cloud phase, droplet or ice crystal size distribution, shape or orientation of ice crystals, cloud overlap, or their combination. The influence of cold clouds is likely second-order due to the much larger typical effective radius of ice crystals than cloud droplets.

In contrast to SW radiation, the models have much smaller LW radiation biases, which is expected due to the prevailing low-level clouds having similar temperatures as the surface. \citeA{roh2021} also found LW biases to be much lower than SW biases in DYAMOND models over the tropical Atlantic Ocean. In ICON, the outgoing LW radiation is overestimated by 5\% (Fig.~\ref{fig:cloud-occurrence}a). This is likely caused by an underestimated total cloud fraction exposing a larger sea surface area to cooling to space, which is typically warmer than the atmospheric temperature at 0--2~km, where most of the clouds are located. In the MERRA-2 and ERA5 reanalyses, the LW biases are also slightly positive, 4 and 5~Wm$^{-2}$, respectively. This is again in line with the underestimated total cloud fraction by about 20\%. However, if the clouds are too thick, as expected from the SW results, this might also provide a compensating effect, in which too small a cloud area is counteracted by greater optical thickness in the LW spectrum, thus reducing the outgoing LW radiation more in thick relative to thinner clouds. For thin clouds, the outgoing TOA LW radiation originates both from the warmer surface (partly blocked by the clouds) and the clouds, whereas for thick clouds, the outgoing TOA LW radiation originates mostly from the colder-than-surface clouds.

In all the subsets (Fig.~\ref{fig:cloud-occurrence}b--g), the same type of biases are observed, namely the outgoing SW radiation is underestimated in ICON and ERA5 and overestimated in MERRA-2, and the outgoing LW radiation is overestimated in all the models. The total cloud fraction is higher by 6\% over the high-latitude SO than the low-latitude SO, and the outgoing SW radiation is much greater by 41~Wm$^{-2}$. ICON has little difference in the total cloud fraction between low- and high-latitude SO, but greater outgoing SW radiation by 14~Wm$^{-2}$ over the high-latitude SO, likely due to thicker clouds and deeper convection in less stable and more cyclonic conditions relative to the low-latitude SO. In contrast, the reanalyses showed both greater total cloud fraction and outgoing SW radiation over the high-latitude SO compared to the low-latitude SO.

@@

In order to examine the potential link in the cloud biases to the local physical conditions, we analyzed about 2300 radiosonde profiles south of 40°S available from the 24 RV \emph{Polarstern} voyages and the MARCUS, NBP1704, TAN1702, and TAN1802 campaigns. Spatially and temporally colocated profiles were taken from ICON and the reanalyses. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), we first remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. The profiles were partitioned into the same subsets as above (Sections~\ref{sec:cloud-occurrence} and \ref{sec:cloud-cover}). We focus on comparing virtual potential temperature ($\theta_v$) and relative humidity, being one of the primary factors affecting shallow convection,the associated low-level cloud formation, and cloud dissipation. The observed and model profiles of virtual potential temperature and relative humidity are shown in Fig.~\ref{fig:potential-temperature}.

Overall, the mean $\theta_v$ is accurate to within 0.5~K in ICON and MERRA-2, except for ICON being colder by up to 2.5~K in the mid-to-high troposphere (less stable) (Fig.~\ref{fig:potential-temperature}a). Larger differences exist, however, in the 40--55°S zone, where ICON is colder by about 5 K at 5~km (Fig.~\ref{fig:potential-temperature}b). In other subsets, the bias is relatively small. MERRA-2 and ERA5 are very close to the observations, possibly due to a high accuracy of assimilation of this quantity. Notably, the variability of virtual potential temperature (as represented by the percentiles) is much smaller in ICON than in the observations. This indicates that the model's internal variability in the lower-tropospheric thermodynamic conditions in the SO is smaller than in reality.

@@

Fig.~\ref{fig:rs-hur-hist} shows RH histograms calculated from the radiosonde observations and equivalent profiles in the models (shown as anomalies relative to the observations), calculated for all, clear, and cloudy bins, based on the lidar observations and the simulated lidar backscatter in the models. Here, we show only the first 2 km to concentrate on the identified cloud biases seen at these heights. We can see several notable features. The models simulate progressively fewer high-RH (\textgreater 90\%) bins above the ground (Fig.~\ref{fig:rs-hur-hist}b--d). This can be related to either ice nucleation happening in the models, which can occur below liquid water saturation, or the grid cell size in the models, which requires lower grid cell average RH than 100\% for saturation to occur in a fraction of the grid cell. The models also tend to simulate more clear bins than observations for RH between 80 and 100\% between the ground and about 1 km (Fig.~\ref{fig:rs-hur-hist}f--h). In the observations, these values of RH are associated with cloudy bins (Fig.~\ref{fig:rs-hur-hist}i). Conversely, the models predominantly associate only RH very close to 100\% with cloudy bins at these heights (Fig.~\ref{fig:rs-hur-hist}j--l). This may be one of the main reasons for the identified cloud or fog biases near the ground. A possible explanation is that cloud droplets are able to form or persist at RH between 90 and 100\% at these heights over the SO. This could be due to abundant sea salt or biogenic aerosol or droplet generation from sea spray in the common high swell and high wind speed conditions over the SO. Stratus fractus or other broken clouds could also lead to less than 100\% RH when averaged over the size of the vertical bins (up to 30 m in some of the radiosonde profiles).

@@

Let us consider the main limitations of the presented results. The spatial coverage of our dataset does not include most parts of the Indian Ocean and Pacific Ocean sectors of the SO. Even though climatological features of the SO are typically relatively uniform zonally, variations exist, such as those related to the Antarctic Peninsula and the southern tip of South America. The voyages were mostly undertaken in the Austral summer months and only rarely in the winter months, due to the poor accessibility of this region during winter. Therefore, our results are likely representative of summer and, to a lesser extent, spring and autumn conditions. Ship access to sea ice covered areas of the SO is also limited. Cloud regimes and phase in the region are seasonally variable \cite{danker2022}.

@@

Supercooled liquid clouds (liquid clouds under subzero temperature) commonly occur over the SO. In our analysis of LWP and IWP, we see that both phases are abundant. Because liquid water droplets are typically smaller and more numerous than ice crystals in cold clouds, they attenuate a greater amount of the lidar radiation. Clouds with a relatively modest optical thickness of 1.7 can attenuate the lidar signal for a detection at 2 km using an instrument with noise properties like the Vaisala CL31 (Section 2.4). While supercooled liquid clouds and their attenuation are accounted for by the lidar simulator, they can strongly attenuate the signal and cause artificially low values of cloud occurrence at higher altitudes. For example, we found that cloud occurrence at 1.5 km is underestimated in ICON and underlying clouds are overestimated. However, this can also mean that clouds at 1.5 km are present in the model but the signal is too attenuated by the lower clouds in the model but not in the observations, where the underlying clouds are not as pronounced.
